Towards a Complete Model of the Negative Selection Process in Humans

### 2023-06-24 manuscript/manuscript.tex
ms updates

The continual influx of new mutations into populations is the ultimate source
of all adaptations, but the vast majority of mutations either do not affect
fitness or are deleterious. Negative selection works to eliminate such
deleterious mutations from the population, which is essential to maintaining
the integrity of information in the genome needed to maintain a lineage.
Mutations under negative selection are kept at low frequencies (Haldane?),
contributing to human traits (XXX) and diseases. As with most selection in the
genome, this process perturbs the allele frequencies of neighboring linked
variation. This reduces genetic variability around conserved segments and
creates a large-scale spatial ``linked selection signal" in diversity alone the
genome. Mutation and negative selection's contribution to this signal is
mediated by the deleterious mutation rate, the strength of selection in
conserved segments, and the spatial distribution of recombination rates and
conserved segments. Since genome-wide recombination maps and putatively
conserved segments are available for many species, researchers have combined
these with population genetic theory to estimate the mutation rate and
selection coefficients of deleterious mutations, as well as the overall
reduction in genetic variability due to the negative selection process.

@@

Our method takes tracks of annotated features (an ``annotation model") that are
\emph{a priori} expected to have similar fitness effects, and estimates the
distribution of fitness effects for each feature type. We consider two classes
of annotation models: (1) CADD score models, which assign a rank-order
pathogenicity score to each basepair, and (2) and more interpretable gene
feature-based models that includes protein coding regions, intron and UTRs, and
PhastCons regions. We include PhastCons regions because there is abundant
evidence of highly-conserved non-coding regions in humans that we would expect
to be under negative selection and impact patterns of linked selection (XXX).
Since our method fits a DFE for each feature type, PhastCons regions that
overlap other gene annotations (e.g. coding sequences) must be assigned to
either of the two feature types. Since this can impact estimates, we fit both
alternatives: a PhastCons Priority model, where features that overlap PhastCons
regions are classified as PhastCons, and a Feature Priority model, where gene
structure features take priority and the PhastCons type catches all
highly-conserved non-genic regions. In total, we fit four annotation models
(CADD 6\%, CADD 8\%, PhastCons Priority, and Feature Priority) to high-coverage
1000 Genome data for three populations: Yoruba (YRI), Han Chinese (CHB), and
European (CEU) ancestries. We assessed and compared models according to how
well they predict the genome-wide patterns of diversity they were fit on. We
use $R_\text{LOO}^2$, which is the proportion of the observed variance in
megabase-scale diversity that is predicted by our model in out-samples (i.e.
whole chromosomes left out during the fitting process).

@@

\begin{figure}[htbp] \centering
    \includegraphics{figures/figure_4_pop_mu_trimmed.pdf} 
    \caption{}
  \label{fig:figure-4}
\end{figure}

@@

>>>>

The continual influx of new mutations into populations is the ultimate source
of all adaptations, but the vast majority of mutations either do not affect
fitness or are deleterious. Negative selection (i.e. purifying selection) works
to eliminate such deleterious mutations from the population, which is essential
to maintaining the integrity of information in the genome needed to maintain a
lineage. Mutations under negative selection are kept at low frequencies
(Haldane?), contributing to human traits (XXX) and diseases. Moreover, the
negative selection process determines sequence conservation across phylogenetic
timescales, which is evident in genome-wide cross-species comparisons
\parencite{Siepel2005-wh}. As with most selection in the genome, the negative
selection process perturbs the allele frequencies of neighboring linked
variation. This reduces genetic variability around conserved segments and
creates a large-scale spatial ``linked selection signal" in diversity alone the
genome. Mutation and negative selection's contribution to this signal is
mediated by the deleterious mutation rate, the strength of selection in
conserved segments, and the spatial distribution of recombination rates and
conserved segments. Since genome-wide recombination maps and putatively
conserved segments are available for many species, researchers have combined
these with population genetic theory to estimate the mutation rate and
selection coefficients of deleterious mutations, as well as the overall
reduction in genetic variability due to the negative selection process.

@@

Our method takes tracks of annotated features (an ``annotation model") that are
\emph{a priori} expected to have similar fitness effects, and estimates the
distribution of fitness effects for each feature type. We consider two classes
of annotation models: (1) CADD score models, which assign a rank-order
pathogenicity score to each basepair, and (2) and more interpretable gene
feature-based models that includes protein coding regions, intron and UTRs, and
PhastCons regions. We include PhastCons regions because there is abundant
evidence of highly-conserved non-coding regions in humans under negative
selection
\parencite{Meader2010-hm,Harmston2013-tt,Katzman2007-gq,Siepel2005-wh}, but
would not be included in gene-based annotation. Since our method fits a DFE for
each feature type, PhastCons regions that overlap other gene annotations (e.g.
coding sequences) must be assigned to either of the two feature types. Since
this can impact estimates, we fit both alternatives: a PhastCons Priority
model, where genic features that overlap PhastCons regions are classified as
PhastCons, and a Feature Priority model, where gene structure features take
priority and the PhastCons type catches all highly-conserved non-genic regions.
In total, we fit four annotation models (CADD 6\%, CADD 8\%, PhastCons
Priority, and Feature Priority) to high-coverage 1000 Genome data for three
populations: Yoruba (YRI), Han Chinese (CHB), and European (CEU) ancestries. We
assess and compare our models according to how well they predict patterns of
diversity on whole chromosomes left-out during the model fitting process. We
use the metric $R_\text{LOO}^2$, which is the proportion of the observed
variance in genomic diversity at the megabase scale predicted by our model. We
experimented with a few smaller spatial scales, but our results were consistent
with previous results suggesting the human linked selection signal fits best at
the megabase scale \parencite{Murphy2022-sj}.

@@

@@

\begin{figure}[htbp] \centering
    \includegraphics[width=\textwidth]{figures/figure_5.pdf} 
    \caption{}
  \label{fig:figure-4}
\end{figure}
